---
title: "reproducible_workflow.Rmd"
author: "Jacobo Dobon"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load libraries
library(tidyverse)
library(here)    # Useful for path reproducibility
library(ggpubr)  # ggplot2 but easier

# Load custom functions
source(here("scripts", "custom_functions.R"))
```

## 1. Data Loading and Self-Checks

We load the data using relative paths with the here package, ensuring that anyone who clones the repository can run the code without changing directories.

```{r load-data}
# Load data
scoring_df <- read_csv(here("data", "BlueberryDiseaseScoring.csv"))
qpcr_df <- read_csv(here("data", "qPCRResult.csv"))

# Self-check: Verify NA in key columns, not present in mine dataset but though
if(any(is.na(scoring_df$Score))) {
  warning("¡Atención! Hay valores perdidos en los Scores.")
} else {
  print("Self-check superado: No hay NAs en la columna Score.")
}
```

## 2. Disease Progression Over Time (Figure 1)

We calculate the mean and standard error of disease progression for all strains.

```{r figure-1-progression, fig.width=10, fig.height=6} 
# Summarize
progression_summary <- scoring_df %>%
  group_by(Strain, Week) %>%
  summarise(
    mean_score = mean(Score, na.rm = TRUE),
    se_score = calc_se(Score),
    .groups = "drop"
  )

# Crear gráfico
fig1 <- ggplot(progression_summary, aes(x = Week, y = mean_score, color = Strain)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_pubr() + # minimal theme for publication
  labs(
    title = "Figure 1: Disease Progression over 9 Weeks",
    x = "Weeks Post-Inoculation",
    y = "Mean Disease Score (0-7)"
  ) +
  theme(legend.position = "right")

# Show plot
fig1
```
*(Note: Since there are 19 strains, this plot will have many lines. If it looks too cluttered in your final visualization, you can add + facet_wrap(~Strain) to the code above).*

## 3. Bacterial Load via qPCR (Figure 2)

We average the technical replicates (leaves/plates) to obtain a single value per plant and compare the strains.

```{r figure-2-qpcr, fig.width=8, fig.height=8}
# Weighted average of bacterial load per plant
qpcr_plant_mean <- qpcr_df %>%
  group_by(Strain, PlantID) %>%
  summarise(mean_logCFU = mean(log10CFUpermg, na.rm = TRUE), .groups = "drop")

# Create Boxplot
fig2 <- ggplot(qpcr_plant_mean, aes(x = reorder(Strain, mean_logCFU), y = mean_logCFU, fill = Strain)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) + # Individual points
  coord_flip() + # Flip axis to read all 19 strains
  theme_pubclean() +
  labs(
    title = "Figure 2: Bacterial Load Across Strains",
    x = "Strain",
    y = "Mean log10(CFU/mg)"
  ) +
  theme(legend.position = "none")

fig2
```

## 4. Correlation: Symptoms vs Bacterial Load (Figure 3)

We cross the data from Week 9 of visual scoring with the qPCR results using an inner_join.

```{r figure-3-correlation, fig.width=8, fig.height=6}
# Extract data from Week 9 and average branches per plant.
week9_scoring <- scoring_df %>%
  filter(Week == 9) %>%
  group_by(Strain, PlantNumber) %>%
  summarise(final_mean_score = mean(Score, na.rm = TRUE), .groups = "drop")

# Join with the qPCR results. 
combined_data <- inner_join(
  week9_scoring, 
  qpcr_plant_mean, 
  by = c("Strain" = "Strain", "PlantNumber" = "PlantID")
)

# Correlation Graph
fig3 <- ggplot(combined_data, aes(x = mean_logCFU, y = final_mean_score)) +
  geom_point(aes(color = Strain), size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed", se = TRUE) +
  theme_pubr() +
  labs(
    title = "Figure 3: Symptom Severity vs Bacterial Load (Week 9)",
    x = "Mean log10(CFU/mg) (qPCR)",
    y = "Final Disease Score (Mean per plant)"
  )

fig3
```